---
title: "Untitled"
author: "Dean de Haas"
date: "27-9-2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

load required libraries

```{r cars}
library(Giotto)
```

set working directory

```{r pressure, echo=FALSE}
my_working_dir = '/mnt/c/Users/deand/OneDrive/Documents/'






```

#set giotto python path
# set python path to your preferred python version path
# set python path to NULL if you want to automatically install (only the 1st time) and use the giotto miniconda environment

```{r pressure, echo=FALSE}
python_path = NULL 
if(is.null(python_path)) {
  installGiottoEnvironment()
}
```

(optional) set Giotto instructions

```{r pressure, echo=FALSE}
 
instrs = createGiottoInstructions(save_plot = TRUE, 
                                  show_plot = FALSE,
                                  save_dir = my_working_dir, 
                                  python_path = python_path)
```

create giotto object from provided paths ####

```{r pressure, echo=FALSE}

expr_path = paste0(my_working_dir, "cortex_svz_expression.txt")
loc_path = paste0(my_working_dir, "cortex_svz_centroids_coord.txt")
meta_path = paste0(my_working_dir, "cortex_svz_centroids_annot.txt")
```

Stitch the multiple field of views together

```{r pressure, echo=FALSE}

## first merge location and additional metadata
SS_locations = data.table::fread(loc_path)
cortex_fields = data.table::fread(meta_path)
SS_loc_annot = data.table::merge.data.table(SS_locations, cortex_fields, by = 'ID')
SS_loc_annot[, ID := factor(ID, levels = paste0('cell_',1:913))]
data.table::setorder(SS_loc_annot, ID)

## create file with offset information
my_offset_file = data.table::data.table(field = c(0, 1, 2, 3, 4, 5, 6),
                                        x_offset = c(0, 1654.97, 1750.75, 1674.35, 675.5, 2048, 675),
                                        y_offset = c(0, 0, 0, 0, -1438.02, -1438.02, 0))

## create a stitch file
stitch_file = stitchFieldCoordinates(location_file = SS_loc_annot,
                                     offset_file = my_offset_file,
                                     cumulate_offset_x = T,
                                     cumulate_offset_y = F,
                                     field_col = 'FOV',
                                     reverse_final_x = F,
                                     reverse_final_y = T)
stitch_file    = stitch_file[,.(ID, X_final, Y_final)]
my_offset_file = my_offset_file[,.(field, x_offset_final, y_offset_final)]
```

create giotto object

```{r pressure, echo=FALSE}
## create Giotto object
SS_seqfish <- createGiottoObject(raw_exprs = expr_path,
                                 spatial_locs = stitch_file,
                                 offset_file = my_offset_file,
                                 instructions = instrs)

## add additional annotation if wanted
SS_seqfish = addCellMetadata(SS_seqfish,
                             new_metadata = cortex_fields,
                             by_column = T,
                             column_cell_ID = 'ID')
```

normalize and take subset of all the cells


```{r pressure, echo=FALSE}
SS_seqfish <- normalizeGiotto(gobject = SS_seqfish, scalefactor = 6000, verbose = T)

cell_metadata = pDataDT(SS_seqfish)
cortex_cell_ids = cell_metadata[FOV %in% 0:3]$cell_ID
SS_seqfish = subsetGiotto(SS_seqfish, cell_ids = cortex_cell_ids)
```

identify spatially informatie genes

```{r pressure, echo=FALSE}
## delaunay network: stats + creation
plotStatDelaunayNetwork(gobject = SS_seqfish, maximum_distance = 400, save_plot = F)
SS_seqfish = createSpatialNetwork(gobject = SS_seqfish, minimum_k = 2, maximum_distance_delaunay = 400)

## create spatial networks based on k and/or distance from centroid
SS_seqfish <- createSpatialNetwork(gobject = SS_seqfish, method = 'kNN', k = 5, name = 'spatial_network')
SS_seqfish <- createSpatialNetwork(gobject = SS_seqfish, method = 'kNN', k = 10, name = 'large_network')
SS_seqfish <- createSpatialNetwork(gobject = SS_seqfish, method = 'kNN', k = 100,
                                   maximum_distance_knn = 200, minimum_k = 2, name = 'distance_network')

## visualize different spatial networks on first field (~ layer 1)
cell_metadata = pDataDT(SS_seqfish)
field1_ids = cell_metadata[FOV == 0]$cell_ID
subSS_seqfish = subsetGiotto(SS_seqfish, cell_ids = field1_ids)

spatPlot(gobject = subSS_seqfish, show_network = T,
         network_color = 'blue', spatial_network_name = 'Delaunay_network',
         point_size = 2.5, cell_color = 'cell_types', 
         save_param = c(save_name = '9_a_spatial_network_delaunay', base_height = 6))




dgft<-SS_seqfish@cell_metadata$cell_ID
# these are the spatially informative genes
km_spatialgenes = binSpect(SS_seqfish)
```

the function from giotto to simulate gene expression levels per cell

```{r pressure, echo=FALSE}
simulateOneGenePatternGiottoObject2 = function(gobject,
                                               pattern_name = 'pattern',
                                               pattern_cell_ids = NULL,
                                               gene_name = NULL,
                                               spatial_prob = 0.95,
                                               gradient_direction = NULL,
                                               show_pattern = TRUE,
                                               pattern_colors = c('in' = 'green', 'out' = 'red'),
                                               ...) {
  
  # data.table variables
  cell_ID = sdimx_y = sdimx = sdimy = NULL
  #print(cell_ID = sdimx_y = sdimx = sdimy = NULL)
  if(is.null(pattern_cell_ids)) {
    stop('pattern_cell_ids can not be NULL \n')
  }
  
  ## create and add annotation for pattern
  cell_meta = pDataDT(gobject)
  cell_meta[, (pattern_name) := ifelse(cell_ID %in% pattern_cell_ids, 'in', 'out')]
  newgobject = addCellMetadata(gobject,
                               new_metadata = cell_meta[,c('cell_ID', pattern_name), with = F],
                               by_column = T,
                               column_cell_ID = 'cell_ID')
  #View(newgobject)
  # show pattern
  if(show_pattern == TRUE) {
    print('we are in')
    spatPlot2D(gobject = newgobject, save_plot = T, cell_color_code = pattern_colors,
               point_size = 2, cell_color = pattern_name)
  }
  
  ## merge cell metadata and cell coordinate data
  cell_meta = pDataDT(newgobject)
  cell_coord = newgobject@spatial_locs
  cell_meta = data.table::merge.data.table(cell_meta, cell_coord, by = 'cell_ID')
  ## get number of cells within pattern
  cell_number = nrow(cell_meta[get(pattern_name) == 'in'])
  #print(cell_number)
  
  ## normalized expression
  expr_data = newgobject@norm_expr
  #View(expr_data)
  #View(newgobject@norm_expr)
  result_list = list()
  
  ## raw expression
  raw_expr_data = newgobject@raw_exprs
  raw_result_list = list()
  
  
  ## create the spatial expression pattern for the specified gene
  # 1. rank all gene values from the cells from high to low
  # 2. move the highest expressing values to the spatial pattern using a probability
  #     - 0.5 is the control = random
  #     - 1 is perfection: all the highest values go to the pattern
  #     - 0.5 to 1 is decreasing noise levels
  
  if(is.null(gene_name)) stop('a gene name needs to be provided')
  
  
  
  # rank genes
  #View(expr_data)
  gene_vector = expr_data[rownames(expr_data) == gene_name, ]
  #View(gene_vector)
  sort_expr_gene = sort(gene_vector, decreasing = T)
  #View(sort_expr_gene)
  # number of cells in and out the pattern
  total_cell_number = length(sort_expr_gene)
  #print('total')
  #print(total_cell_number)
  #print(sort_expr_gene)
  remaining_cell_number = total_cell_number - cell_number
  
  # calculate outside probability
  outside_prob = 1 - spatial_prob
  #print(outside_prob)
  #print('rep1')
  #print(spatial_prob)
  #print(remaining_cell_number)
  prob_vector = c(rep(spatial_prob, cell_number), rep(outside_prob, remaining_cell_number))
  #print('rep2')
  # first get the 'in' pattern sample values randomly
  sample_values = sample(sort_expr_gene, replace = F, size = cell_number, prob = prob_vector)
  
  # then take the remaining 'out' pattern values randomly
  remain_values = sort_expr_gene[!names(sort_expr_gene) %in% names(sample_values)]
  remain_values = sample(remain_values, size = length(remain_values))
  
  
  
  ## A. within pattern ##
  # ------------------- #
  in_cell_meta = cell_meta[get(pattern_name) == 'in']
  
  # if gradient is wanted
  # does not work with 0.5!! is not random!!
  if(!is.null(gradient_direction)) {
    # sort in_ids according to x, y or  xy coordinates to create gradient
    in_cell_meta[, sdimx_y := abs(sdimx)+ abs(sdimy)]
    # order according to gradient direction
    in_cell_meta = in_cell_meta[order(get(gradient_direction))]
  }
  in_ids = in_cell_meta$cell_ID
  
  # preparation for raw matrix
  sample_values_id_vector = names(sample_values)
  names(sample_values_id_vector) = in_ids
  
  
  ## B. outside pattern ##
  # -------------------- #
  out_ids = cell_meta[get(pattern_name) == 'out']$cell_ID
  
  # preparation for raw matrix
  remain_values_id_vector = names(remain_values)
  names(remain_values_id_vector) = out_ids
  
  ## raw matrix
  # swap the cell ids #
  raw_gene_vector = raw_expr_data[rownames(raw_expr_data) == gene_name,]
  
  raw_new_sample_vector = raw_gene_vector[sample_values_id_vector]
  names(raw_new_sample_vector) = names(sample_values_id_vector)
  
  raw_new_remain_vector = raw_gene_vector[remain_values_id_vector]
  names(raw_new_remain_vector) = names(remain_values_id_vector)
  
  new_sim_raw_values = c(raw_new_sample_vector, raw_new_remain_vector)
  new_sim_raw_values = new_sim_raw_values[names(raw_gene_vector)]
  
  # change the original matrices
  raw_expr_data[rownames(raw_expr_data) == gene_name,] = new_sim_raw_values

  exprrr=new_sim_raw_values
  newgobject@raw_exprs = raw_expr_data
  
  # recalculate normalized values
  newgobject <- normalizeGiotto(gobject = newgobject, ...)
  newgobject <- addStatistics(gobject = newgobject)
  
  return(new_sim_raw_values)
  
}

```

```{r pressure, echo=FALSE}
#below selected cells will have higher gene expressions
selected_cells = cell_metadata[FOV %in% 2:3]$cell_ID
#slect the top spatially informative genes
top_spatially_informative_genes<-c('A'=head(km_spatialgenes$genes,42))
# assign higher expression levels to these genes in the selected cella
simgene=simulateOneGenePatternGiottoObject2(SS_seqfish,pattern_name = "in", pattern_cell_ids = selected_cells, gene_name= top_spatially_informative_genes[1],show_pattern = T)
sim_gene_matrix<-as.matrix(simgene)
sim_gene_dataframe<-data.frame(simgene)
simulated_gene_dataset<-data.frame('cells'=rownames(sim_gene_dataframe))
for (gene in top_spatially_informative_genes) {
  simgene=data.frame(simulateOneGenePatternGiottoObject2(SS_seqfish,pattern_name = "in", pattern_cell_ids = selected_cells, gene_name= gene,show_pattern = T))
  colnames(simgene)<-gene
  simulated_gene_dataset<-cbind(simulated_gene_dataset,simgene)
}
#write the dataframe with new expression levels to output
frame<-simulated_gene_dataset[,-c(1)]
write.table(frame,'/mnt/c/Users/deand/OneDrive/Documents/simulldata2.txt')
```

```{r pressure, echo=FALSE}

simulateOneGenePatternGiottoObject3 = function(gobject,
                                               pattern_name = 'pattern',
                                               pattern_cell_ids = NULL,
                                               gene_name = NULL,
                                               spatial_prob = 0.95,
                                               gradient_direction = NULL,
                                               show_pattern = TRUE,
                                               pattern_colors = c('in' = 'green', 'out' = 'red'),
                                               ...) {
  
  # data.table variables
  cell_ID = sdimx_y = sdimx = sdimy = NULL
  #print(cell_ID = sdimx_y = sdimx = sdimy = NULL)
  if(is.null(pattern_cell_ids)) {
    stop('pattern_cell_ids can not be NULL \n')
  }
  
  ## create and add annotation for pattern
  cell_meta = pDataDT(gobject)
  cell_meta[, (pattern_name) := ifelse(cell_ID %in% pattern_cell_ids, 'in', 'out')]
  newgobject = addCellMetadata(gobject,
                               new_metadata = cell_meta[,c('cell_ID', pattern_name), with = F],
                               by_column = T,
                               column_cell_ID = 'cell_ID')
  #View(newgobject)
  # show pattern
  if(show_pattern == TRUE) {
    print('we are in')
    spatPlot2D(gobject = newgobject, save_plot = T, cell_color_code = pattern_colors,
               point_size = 2, cell_color = pattern_name)
  }
  
  ## merge cell metadata and cell coordinate data
  cell_meta = pDataDT(newgobject)
  cell_coord = newgobject@spatial_locs
  cell_meta = data.table::merge.data.table(cell_meta, cell_coord, by = 'cell_ID')
  ## get number of cells within pattern
  cell_number = nrow(cell_meta[get(pattern_name) == 'in'])
  #print(cell_number)
  
  ## normalized expression
  expr_data = newgobject@norm_expr
  #View(expr_data)
  #View(newgobject@norm_expr)
  result_list = list()
  
  ## raw expression
  raw_expr_data = newgobject@raw_exprs
  raw_result_list = list()
  
  
  ## create the spatial expression pattern for the specified gene
  # 1. rank all gene values from the cells from high to low
  # 2. move the highest expressing values to the spatial pattern using a probability
  #     - 0.5 is the control = random
  #     - 1 is perfection: all the highest values go to the pattern
  #     - 0.5 to 1 is decreasing noise levels
  
  if(is.null(gene_name)) stop('a gene name needs to be provided')
  
  
  
  # rank genes
  #View(expr_data)
  gene_vector = expr_data[rownames(expr_data) == gene_name, ]
  #View(gene_vector)
  sort_expr_gene = sort(gene_vector, decreasing = T)
  #View(sort_expr_gene)
  # number of cells in and out the pattern
  total_cell_number = length(sort_expr_gene)
  #print('total')
  #print(total_cell_number)
  #print(sort_expr_gene)
  remaining_cell_number = total_cell_number - cell_number
  
  # calculate outside probability
  outside_prob = 1 - spatial_prob
  #print(outside_prob)
  #print('rep1')
  #print(spatial_prob)
  #print(remaining_cell_number)
  prob_vector = c(rep(spatial_prob, cell_number), rep(outside_prob, remaining_cell_number))
  #print('rep2')
  # first get the 'in' pattern sample values randomly
  sample_values = sample(sort_expr_gene, replace = F, size = cell_number, prob = prob_vector)
  
  # then take the remaining 'out' pattern values randomly
  remain_values = sort_expr_gene[!names(sort_expr_gene) %in% names(sample_values)]
  remain_values = sample(remain_values, size = length(remain_values))
  
  
  
  ## A. within pattern ##
  # ------------------- #
  in_cell_meta = cell_meta[get(pattern_name) == 'in']
  
  # if gradient is wanted
  # does not work with 0.5!! is not random!!
  if(!is.null(gradient_direction)) {
    # sort in_ids according to x, y or  xy coordinates to create gradient
    in_cell_meta[, sdimx_y := abs(sdimx)+ abs(sdimy)]
    # order according to gradient direction
    in_cell_meta = in_cell_meta[order(get(gradient_direction))]
  }
  in_ids = in_cell_meta$cell_ID
  
  # preparation for raw matrix
  sample_values_id_vector = names(sample_values)
  names(sample_values_id_vector) = in_ids
  
  
  ## B. outside pattern ##
  # -------------------- #
  out_ids = cell_meta[get(pattern_name) == 'out']$cell_ID
  
  # preparation for raw matrix
  remain_values_id_vector = names(remain_values)
  names(remain_values_id_vector) = out_ids
  
  ## raw matrix
  # swap the cell ids #
  raw_gene_vector = raw_expr_data[rownames(raw_expr_data) == gene_name,]
  
  raw_new_sample_vector = raw_gene_vector[sample_values_id_vector]
  names(raw_new_sample_vector) = names(sample_values_id_vector)
  
  raw_new_remain_vector = raw_gene_vector[remain_values_id_vector]
  names(raw_new_remain_vector) = names(remain_values_id_vector)
  
  new_sim_raw_values = c(raw_new_sample_vector, raw_new_remain_vector)
  new_sim_raw_values = new_sim_raw_values[names(raw_gene_vector)]
  
  # change the original matrices
  raw_expr_data[rownames(raw_expr_data) == gene_name,] = new_sim_raw_values
  
  exprrr=new_sim_raw_values
  newgobject@raw_exprs = raw_expr_data
  
  # recalculate normalized values
  newgobject <- normalizeGiotto(gobject = newgobject, ...)
  newgobject <- addStatistics(gobject = newgobject)
  
  return(newgobject)
  
}


```

```{r pressure, echo=FALSE}

frame<-simulated_gene_dataset[,-c(1)]
v<-rowSums(frame)
ddf<-data.frame(SS_seqfish@spatial_locs$sdimx,SS_seqfish@spatial_locs$sdimy,sim_gene_matrix)

library(ggplot2)
flitz=simulateOneGenePatternGiottoObject3(SS_seqfish,pattern_name = "in", pattern_cell_ids = pszt, gene_name= df[1],show_pattern = T)
View(flitz@cell_metadata)
zz<-flitz@cell_metadata[,c(1,4)]
colnames(zz)[2]<-'group'
ggplot(ddf,aes(SS_seqfish@spatial_locs$sdimx,SS_seqfish@spatial_locs$sdimy, colour = zz$group ))+geom_point()

locs<-data.frame('cells'=zz$cell_ID,'X'=SS_seqfish@spatial_locs$sdimx,'Y'=SS_seqfish@spatial_locs$sdimy,'group'=zz$group)
write.table(locs,'/mnt/c/Users/deand/OneDrive/Documents/simulldata_locs.txt')


ggplot(ddf,aes(SS_seqfish@spatial_locs$sdimx,SS_seqfish@spatial_locs$sdimy, color = v ))+geom_point()

View(as.data.frame(rowSums(ddf)))


```








